---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
    pandoc_args: ["+RTS", "-K64m", "-RTS"]
---

      
## Loading and preprocessing the data
```{r, echo=TRUE, results='hide'}
data <- read.csv("activity.csv")
```


## What is mean total number of steps taken per day?
The ddply function is used to calculate the number of steps per day.

```{r, echo=TRUE}
library(plyr)
sumSteps <- ddply(data, "date", summarise, steps = sum(steps, na.rm=TRUE))
hist(sumSteps$steps, breaks=10, col='lightblue', main='Histogram of the number of steps', xlim=c(0,25000), xlab='Number of steps')
summary(sumSteps$steps)
```


## What is the average daily activity pattern?
First, the dates need to be converted into days of the week. A column with this information is added to the dataframe.
Then the average per day is calculated using the ddply function. The results are plotted and the maximum is calculated.

```{r, echo=TRUE}
data$day <- factor(weekdays(as.Date(data$date)))
data$day <- factor(data$day, levels= c("Sunday", "Monday", 
    "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
avgByInt <- ddply(data, .(interval), summarise, steps = mean(steps, na.rm=TRUE))
invisible(avgByInt[order(data$day), ])
plot(avgByInt$steps, type='l', xlab='Interval', ylab='Number of steps')
avgByInt[which.max(avgByInt$steps),]
```


## Imputing missing values
The missing values are replaced by the mean of that particular interval for the specific day.

```{r, echo=TRUE}
sum(is.na(data$steps))
idx <- is.na(data$steps)
modData <- data
avgByDay <- ddply(data, .(day, interval), summarise, steps = mean(steps, na.rm=TRUE))
for (i in 1:length(idx)) {
      if (idx[i]) {
            modData$steps[i] <- avgByDay$steps[which(avgByDay$day==data$day[i] & avgByDay$interval==data$interval[i])]
      }
}
```

Now, the same methods as before can be used to make a histogram and calculate the mean and median.

```{r, echo=TRUE}
sumStepsMod <- ddply(modData, "date", summarise, steps = sum(steps, na.rm=TRUE))
hist(sumStepsMod$steps, breaks=10, col='lightblue', main='Histogram of the number of steps', xlim=c(0,25000), xlab='Number of steps')
summary(sumStepsMod$steps)
```
Filling in the missing values slightly changes the values of the minimum, mean, and median of the number of steps.

## Are there differences in activity patterns between weekdays and weekends?
First, a new factor variable is created to indicate if a day is a weekday or in the weekend.
```{r, echo=TRUE}
modData$weekend <- factor(modData$day %in% c("Saturday", "Sunday"))
modData$weekend <- revalue(modData$weekend, c("FALSE"="weekday", "TRUE"="weekend"))
```

The mean of the steps is calculated and the plot is made.
```{r, echo=TRUE}
avgByWeekend <- ddply(modData, .(weekend, interval), summarise, steps = mean(steps))
library(ggplot2)
qplot(interval, steps, data=avgByWeekend, facets=weekend~., geom="line", xlab="Interval", ylab="Number of steps")
```